\section*{Working Findings Log}

\subsection*{2026-06-27 Codex Loop}

\textbf{Standing transcript check.} Re-read the priority JSON transcript artifacts for government surveillance and UBI under seed~17. The previous transcript-review grades still hold. Government surveillance mixed-language remains the best hook: B-Indonesian Turn~4 introduces surveillance for \textit{tujuan politik atau ekonomi} and proposes lawful purpose limitation plus independent oversight, while same-English B Turn~4 moves to encryption/anonymization and technical feasibility. UBI remains weak as an elicitation case: B-Indonesian repeats the inflation-management frame in Turn~2 and introduces only generic conditionality in Turn~4.

\textbf{New-run grading.} No new empirical run artifacts were produced in this loop, so no new runs were graded. The next loop should grade any artifacts created by the new Modal batch entrypoints before adding results to the paper.

\textbf{Implementation state.} `code/modal_bivad_runner.py` now has explicit batch entrypoints for: (1) engagement-prompt reruns of UBI and government surveillance, (2) Arabic/Hindi/French/Chinese expansion on UBI and government surveillance, and (3) culturally loaded Indonesian/Spanish topic scans. Artifacts now include a top-level `target_language` field and batch-tagged run IDs/manifests for future runs.

\textbf{Blocked empirical items.} The language expansion, culturally loaded topic scan, and steering retry remain empirically pending because no Modal GPU jobs were launched in this loop. Smallest next action is:
\[
\texttt{python3 -m modal run code/modal\_bivad\_runner.py::rerun\_top\_divergence}
\]
Then manually grade the four transcripts in `code/bivad-evidence-audit/transcript_review.md` before running the larger expansion grid.
