\documentclass[10pt]{article}
\usepackage[margin=0.8in]{geometry}
\usepackage{array}
\usepackage{longtable}
\usepackage{url}
\usepackage{hyperref}
\hypersetup{colorlinks=false,pdfborder={0 0 0}}

\title{B5 NegotiationArena Cross-Lingual Contact: Seed-101 OpenAI Evidence}
\author{negotiation\_arena\_crosslingual}
\date{2026-07-13}

\newcommand{\artifact}[1]{\path{#1}}

\begin{document}
\maketitle

\section*{Answer}

For the current B5 NegotiationArena evidence, the answer is a constrained null:
the seed-101 OpenAI \texttt{gpt-5.4-mini-2026-03-17} runs do not support a
stable higher-resource language payoff premium after counterbalancing. All
buy/sell C2 rows reached deals and parsed offers, but the payoff differences
were role-sensitive and two C2 rows failed strict assigned-channel compliance
because visible dialogue converged to Chinese. These are OpenAI benchmark
override results, not Qwen3 evidence.

H1 is not supported. EN-ID showed EN minus ID payoff asymmetry of +36 when EN
was buyer, but 0 when EN was seller. EN-ZH flipped sign, -10 then +8. ZH-ID was
-20 then 0 for ZH minus ID. H2 therefore fails as a language-premium claim:
the counterbalances do not show a stable same-sign premium. H3 is weak and
mostly first-offer/seller driven rather than higher-resource-channel driven.
Final prices were 0--10 ZUP away from the first offer, and the first offer was
always made by the seller. H4 is supported only as a convergence observation:
C3 converged to EN for EN-ID, to ZH for EN-ZH, and to ZH for ZH-ID, with no
detected off-pair English leakage in ZH-ID. H5 is the best current summary:
the capability floor passed, C2 outcomes were not worse than baselines, and no
counterbalanced language premium remains defensible from this evidence.

\section*{Design and Gates}

Language is the visible interaction-output channel. Rules and private state
remain in English. The active language pairs are EN-ID, EN-ZH, and ZH-ID. C0
and C1 are monolingual output-channel baselines; C2 assigns each role one
language and counterbalances buyer/seller role; C3 permits either language in
the active pair. Source bring-up used the local NegotiationArena checkout at
\artifact{external/NegotiationArena}, branch \texttt{paper\_experiment\_code},
commit \texttt{d35a7a3aa0d94c2d49f1d6ac13c5f931851abf12}.

The G2 capability gate passed. C0 EN buy/sell reached a deal with offer parse
rate 1.0 and channel compliance 1.0 at
\artifact{artifacts/results/baseline_c0_buy_sell_en_seed001.openai_benchmark.metrics.json}.
C1 ID buy/sell likewise reached a deal with offer parse rate 1.0 and channel
compliance 1.0 at
\artifact{artifacts/results/baseline_c1_buy_sell_id_seed001.metrics.json}. C1
ZH buy/sell reached a deal with offer parse rate 1.0 and channel compliance 1.0
at \artifact{artifacts/results/pair_en_zh_c1_buy_sell_seed101.metrics.json}.
The gate artifact is \artifact{artifacts/results/g2_capability_floor.json}.

\section*{Evidence Table}

\begin{center}
\small
\begin{tabular}{|p{0.12\linewidth}|p{0.16\linewidth}|p{0.11\linewidth}|p{0.11\linewidth}|p{0.12\linewidth}|p{0.12\linewidth}|p{0.13\linewidth}|}
\hline
Pair/condition & Assignment & Final & Payoffs & Lx-Ly payoff & Channel compliance & Interpretation \\
\hline
EN-ID C2 & buyer EN, seller ID & price 52 & buyer 48, seller 12 & +36 & 1.00 & EN advantage only with EN as buyer \\
\hline
EN-ID C2 & buyer ID, seller EN & price 70 & buyer 30, seller 30 & 0 & 1.00 & counterbalance removes EN premium \\
\hline
EN-ZH C2 & buyer EN, seller ZH & price 75 & buyer 25, seller 35 & -10 & 0.50 & Chinese convergence; no EN premium \\
\hline
EN-ZH C2 & buyer ZH, seller EN & price 74 & buyer 26, seller 34 & +8 & 0.667 & sign flip; compliance limited \\
\hline
ZH-ID C2 & buyer ZH, seller ID & price 80 & buyer 20, seller 40 & -20 & 1.00 & no ZH premium \\
\hline
ZH-ID C2 & buyer ID, seller ZH & price 70 & buyer 30, seller 30 & 0 & 0.50 & Chinese convergence; no stable premium \\
\hline
EN-ID C3 & bilingual EN/ID & price 50 & buyer 50, seller 10 & n/a & 1.00 & converged to EN; off-pair 0 \\
\hline
EN-ZH C3 & bilingual EN/ZH & price 50 & buyer 50, seller 10 & n/a & 1.00 & converged to ZH; off-pair 0 \\
\hline
ZH-ID C3 & bilingual ZH/ID & price 60 & buyer 40, seller 20 & n/a & 1.00 & converged to ZH; EN leakage 0 \\
\hline
\end{tabular}
\end{center}

The C2 buy/sell rows cite, in order:
\artifact{artifacts/results/pair_en_id_c2_buyer_lx_seller_ly_buy_sell_seed101.metrics.json},
\artifact{artifacts/results/pair_en_id_c2_buyer_ly_seller_lx_buy_sell_seed101.metrics.json},
\artifact{artifacts/results/pair_en_zh_c2_buyer_lx_seller_ly_buy_sell_seed101.metrics.json},
\artifact{artifacts/results/pair_en_zh_c2_buyer_ly_seller_lx_buy_sell_seed101.metrics.json},
\artifact{artifacts/results/pair_zh_id_c2_buyer_lx_seller_ly_buy_sell_seed101.metrics.json}, and
\artifact{artifacts/results/pair_zh_id_c2_buyer_ly_seller_lx_buy_sell_seed101.metrics.json}.
The C3 rows cite
\artifact{artifacts/results/pair_en_id_c3_buy_sell_seed101.metrics.json},
\artifact{artifacts/results/pair_en_zh_c3_buy_sell_seed101.metrics.json}, and
\artifact{artifacts/results/pair_zh_id_c3_buy_sell_seed101.metrics.json}.

\section*{First-Offer Anchoring}

The buy/sell first-offer metric is final price minus first offer price. The
EN-ID rows were -8 when the ID seller opened at 60 and -10 when the EN seller
opened at 80. EN-ZH was -5 when the ZH seller opened at 80 and -6 when the EN
seller opened at 80. ZH-ID was 0 when the ID seller opened at 80 and -10 when
the ZH seller opened at 80. This is not a clean higher-resource anchoring
effect; it is more consistent with seller-first-offer anchoring plus role and
compliance effects. The one completed resource-exchange C2 row,
\artifact{artifacts/results/pair_en_id_c2_buyer_lx_seller_ly_resource_exchange_seed101.metrics.json},
ended exactly at the EN agent's first allocation offer, but resource exchange
is not yet counterbalanced.

\section*{Representative Interaction Excerpts}

EN-ID C2 buyer-EN/seller-ID:
the ID seller opened, ``Saya menawarkan X dengan harga 60 ZUP,'' the EN buyer
countered, ``I can do 50 ZUP for X,'' and the ID seller accepted 52 ZUP. Full
transcript:
\artifact{artifacts/transcripts/pair_en_id_c2_buyer_lx_seller_ly_buy_sell_seed101.json}.

EN-ZH C2 buyer-EN/seller-ZH:
the seller opened in Chinese with a translated offer of 1 X for 80 ZUP, and the
buyer assigned to English also replied in Chinese with a 70 ZUP counteroffer.
This explains the 0.50 assigned-channel compliance despite a valid deal.
Transcript:
\artifact{artifacts/transcripts/pair_en_zh_c2_buyer_lx_seller_ly_buy_sell_seed101.json}.

ZH-ID C2 buyer-ID/seller-ZH:
the ZH seller opened at 80 ZUP; the ID-assigned buyer replied in Chinese with
60 ZUP and later accepted 70 ZUP. This row is useful evidence of channel drift,
not a clean assigned-language treatment. Transcript:
\artifact{artifacts/transcripts/pair_zh_id_c2_buyer_ly_seller_lx_buy_sell_seed101.json}.

C3 free choice:
EN-ID used English throughout at
\artifact{artifacts/transcripts/pair_en_id_c3_buy_sell_seed101.json}; EN-ZH
used Chinese throughout at
\artifact{artifacts/transcripts/pair_en_zh_c3_buy_sell_seed101.json}; ZH-ID
used Chinese throughout, with no detected English leakage, at
\artifact{artifacts/transcripts/pair_zh_id_c3_buy_sell_seed101.json}.

\section*{Limitations and Next Work}

This report is intentionally narrow. It is a seed-101 OpenAI benchmark evidence
report and should not be generalized as a Qwen3 matrix result. The buy/sell
matrix is counterbalanced for all active pairs, but it has only one seed per
row. EN-ZH and one ZH-ID C2 row are compliance-limited because agents converged
to Chinese despite assigned channels. Resource exchange is executable for all
active pairs, but only the EN-ID agent-a-EN/agent-b-ID row has been run; the
next exact resource-exchange commands are listed in
\artifact{reports/findings.md}. More seeds and completed resource-exchange
counterbalances are needed before estimating effect sizes.

\section*{Conclusion}

The current answer to the headline question is no: in this evidence bundle,
the higher-resource visible language channel does not capture a robust
counterbalanced payoff premium. The clearest positive finding is process-level:
free-choice language selection can converge strongly to one visible language,
and in these seed-101 runs it converged to EN for EN-ID but to ZH for EN-ZH and
ZH-ID.

\end{document}
